R Course Syllabus

### 2023-04-18 Draft Syllabus.Rmd
Continuing lesson 5

## Lesson 5 - Conditionals and loops {.tabset .tabset-fade .tabset-pills}

@@

1. If/then conditionals and their use
    + When to use them
2. While loops
    + When to use
3. For loops
    + When to use

@@

#### While loops

@@

```{r}
i <- 0
total <- 0
while (i <= 100) {
  total <- total + i
  i <- i + 1
}
total
```

>>>>

## Lesson 5 - Control Flow {.tabset .tabset-fade .tabset-pills}

@@

1. If/then conditionals and their use
2. While loops
3. For loops
4. A brief discussion of map

@@

#### **While loops**

@@

```{r l5_2_1}
i <- 0
total <- 0
while (i <= 100) {
  total <- total + i
  i <- i + 1
}
total
i
```

Outside of the loop we defined the variables to i and total to track the iteration of the loop that we are on and the total respectively. The while loop is made up of the statement `while`, the condition wrapped in parentheses, and the curly braces `{}` which enclose the code to be executed while the condition is TRUE. As soon as the condition becomes false--that is, i > 100, then the while loop stops executing.

When working with `while` loops, be careful not to create an infinite loop. If you create a condition which is always true then your code will run forever and you will have to manually break the program.

```{r l5_2_2}
#' i <- 1
#' while (i > 0) {
#'  i <- i + 1
#' }
#' The code above will run forever because i will always be greater than 0
```

#### **For loops**

It is often the case that you will want to repeat the same code when performing data analysis. One way to do this is with the for loop.

```{r l5_3_1}
for (i in c(0, 1)) {
  print(i)
}
```

We begin with the `for` statement, then in parentheses we set the range or values that we want `for` to operate over, and then in the curly braces `{}` we place the code that we would like repeated. People often create for loops using the `seq()` function:

```{r l5_3_2}
sum <- 0
for (i in seq(from = 2, to = 10, by = 2)) {
  sum <- sum + i
}
sum
```
But you can loop over any vector.

```{r l5_3_3}
names <- c("Reggie", "Gilliard", "Jr.")
for (i in names) {
  print(i)
}
```

#### **map**

Surprisingly, `for` loops are not that common in R. Most often, people use a family functions called the apply functions. These functions loop over lists, vectors, matrices, and data frames and return objects of the same type. But dplyr has its own versions of these functions. I'll only introduce you to one here, `map`.

With the `map` function you provide a vector of values that you would like to loop over, and then a function to apply to those values. You can provide the function to map in 3 different ways.

```{r l5_4_1}
# Defining a function that will multiply a value by 10
mult10 <- function(x) {
  return(x*10)
}

# Use the tilde to create an anonymous funciton
map(1:10, ~ .x * 10)

# Defining an anonymous function inline
map(1:10, function(x) x*10)

# Using an extant function
map(1:10, mult10)

# Each of these is equivalent to running
list(1 * 10, 2 * 10, 3 * 10, 4 * 10, 5 * 10, 6 * 10, 7 * 10, 8 * 10, 9 * 10, 10 * 10)
```

Even in these examples it is clear that functions like `map()` (formally called functionals) make repeating tasks much more efficient. You are much more likely to see functions like map, or lapply when working with code in R, so as you continue your coding journey it is worthwhile to investigate them more deeply.

### **Helpful Links**

* [Codecademy Lesson](https://www.codecademy.com/courses/learn-r/lessons/intro-visualization-ggplot2-r/exercises/review)
* [W3 Schools](https://www.w3schools.com/r/r_if_else.asp)
* [Functionals](https://adv-r.hadley.nz/functionals.html#map)
